Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-07-03 docs/paper.Rmd
write conceptual framework & hypothesis

>>>>

\section{Conceptual Framework and Hypotheses}

This study builds upon the foundational framework of the quantity–quality (Q–Q) trade-off, as articulated in \citet{becker1960economic} seminal work and later refined by \citet{becker1973interaction}. Within this theoretical model, fertility and child quality are jointly determined under a full-income constraint, such that any increase in the number of children—absent a commensurate expansion in resources—necessitates a reduction in the average investment per child. While prior empirical applications of this framework have largely centered on educational attainment, child quality may also be understood along biological dimensions, especially along nutritional status, which emerges earlier in the life course and may be less subject to later remediation.

The current analysis focuses exclusively on early-life nutrition as a proxy for child quality. Nutrition, unlike cognitive or educational outcomes, is characterized by low substitutability and a narrow temporal window during which investments are most productive. The trade-off is thus expected to manifest strongly in settings where fertility rises exogenously while household resources remain constrained. In such environments, families facing an unexpected increase in sibship size may be forced to reallocate scarce food, time, and health inputs across a larger number of dependents, potentially resulting in lower nutritional attainment for each child.

To formalize this relationship, consider a household with total income \( I \), choosing nutritional investments \(\kappa_j\) for each of \( n \) children, along with consumption \( Z \) unrelated to child quality:

\[
I = \sum_{j=1}^{n} \kappa_j + Z.
\]

In the presence of a fertility shock, holding \( I \) constant, the average \(\kappa_j\) must fall unless offset by changes in parental behavior or external transfers. If nutritional investments per child decline, and if such inputs are critical for growth, observable deficits in height-for-age (HAZ) or weight-for-height (WHZ) may emerge.
The first hypothesis emerging from this framework is that an exogenous increase in

fertility leads to a reduction in per-child nutritional investment, particularly in low-income, urban households where slack in the family budget is limited. A second, related hypothesis posits that this reduction in inputs translates into measurable nutritional deficits among young children, specifically in the form of lower anthropometric outcomes such as height-for-age (HAZ) and weight-for-height (WHZ) z-scores. These two hypotheses—concerning, respectively, the behavioral reallocation of resources and its biological consequence—jointly seek to identify a causal link between fertility shocks and early-life deprivation.

### 2025-07-03 docs/paper.tex
write conceptual framework & hypothesis

>>>>

\section{Conceptual Framework and Hypotheses}

This study builds upon the foundational framework of the
quantity--quality (Q--Q) trade-off, as articulated in
\citet{becker1960economic} seminal work and later refined by
\citet{becker1973interaction}. Within this theoretical model, fertility
and child quality are jointly determined under a full-income constraint,
such that any increase in the number of children---absent a commensurate
expansion in resources---necessitates a reduction in the average
investment per child. While prior empirical applications of this
framework have largely centered on educational attainment, child quality
may also be understood along biological dimensions, especially along
nutritional status, which emerges earlier in the life course and may be
less subject to later remediation.

The current analysis focuses exclusively on early-life nutrition as a
proxy for child quality. Nutrition, unlike cognitive or educational
outcomes, is characterized by low substitutability and a narrow temporal
window during which investments are most productive. The trade-off is
thus expected to manifest strongly in settings where fertility rises
exogenously while household resources remain constrained. In such
environments, families facing an unexpected increase in sibship size may
be forced to reallocate scarce food, time, and health inputs across a
larger number of dependents, potentially resulting in lower nutritional
attainment for each child.

To formalize this relationship, consider a household with total income
\(I\), choosing nutritional investments \(\kappa_j\) for each of \(n\)
children, along with consumption \(Z\) unrelated to child quality:

\[
I = \sum_{j=1}^{n} \kappa_j + Z.
\]

In the presence of a fertility shock, holding \(I\) constant, the
average \(\kappa_j\) must fall unless offset by changes in parental
behavior or external transfers. If nutritional investments per child
decline, and if such inputs are critical for growth, observable deficits
in height-for-age (HAZ) or weight-for-height (WHZ) may emerge. The first
hypothesis emerging from this framework is that an exogenous increase in

fertility leads to a reduction in per-child nutritional investment,
particularly in low-income, urban households where slack in the family
budget is limited. A second, related hypothesis posits that this
reduction in inputs translates into measurable nutritional deficits
among young children, specifically in the form of lower anthropometric
outcomes such as height-for-age (HAZ) and weight-for-height (WHZ)
z-scores. These two hypotheses---concerning, respectively, the
behavioral reallocation of resources and its biological
consequence---jointly seek to identify a causal link between fertility
shocks and early-life deprivation.
